Spitzer, Gaia and the Potential of the Milky Way

### 2013-07-31 text/paper1/paper.tex
revisions

The Milky Way is the best candidate  for dissecting a dark matter halo in great detail.
Our proximity means that we can use individual stars as tracer populations and hence build much larger samples that probe deeper into the halo than the globular cluster
and planetary nebula studies of external galaxies \citep[e.g.][]{mendez01,cote03}.
Moreover, our internal perspective allows us to build a three-dimensional picture of the mass distribution rather than measuring its two-dimensional projection.
%Hence we can hope to reconstruct not only the density, but also the shape and orientation of the dark matter halo as function of radius well beyond the disk.
For example, to date thousands of blue-horizontal-branch stars selected from the Sloan Digital Sky Survey  have been used to probe mass out to tens of
kpc \citep[SDSS, see][]{deason12a,kafle12}, and estimates with combined tracers extend to 150kpc \citep{deason12b}.
\citet{loebman12} demonstrates how even larger stellar samples from current (i.e. SDSS) and future  (Gaia and LSST) stellar surveys will also be sensitive to any
asphericity of the halo.

@@

In this paper we re-examine and update  another approach to using tidal debris as a potential measure \citep[originially proposed by][]{johnston99a}  in the context of current and near-future observational capabilities, and illustrates its power with the example of Sgr.
In Section \ref{context.sec} we outline the observational prospects and Sgr properties that motivated this re-examination.
In Section \ref{measure.sec} we present the updated potential measure and test it with synthetic observations of simulated Sgr debris.
In Section \ref{summary.sec} we summarize our results and outline prospects for combining constraints from Sgr with those from other debris systems.

@@

\section{Context and motivation}
The method presented in Section \ref{sec:method} takes advantage of
three distinct developments in the observations of tidal streams: (i)
the demonstration of a technique for deriving distances to individual
RR Lyrae stars with 2\% accuracies (Section \ref{sec:spitzer}); (ii)
the prospect of proper motion measurements of the same stars with
$\sim$10~$\mu$as/yr precision (Section \ref{sec:gaia}); and (iii) the
tracing of debris associated with Sgr around the entire Galaxy
(Section \ref{sec:sgr})

@@

These studies suggest that observations of RR Lyrae variables by
NASA's Spitzer mission could provide us with a stellar distance
indicator of unprecedented accuracy and precision.
\citet{madore12}
For example, using 5 stars with trigonometric parallaxes
  measured by Hubble \citep{benedict11}, \citet{madore12} measured  
  the dispersion in the
mid-IR PL relation -- at Spitzer wavelengths --  to be of order
$\pm$0.03~mag. This implies that it is possible to determine
individual distances that are good to better than $\pm 2\%$ for RR
Lyrae stars out to $\sim$60~kpc (the limit of Spitzer's ability to
detect and measure RR Lyrae variables, for modest integration
times). Distance measurements of Blue Horizontal Branch and Blue Straggler stars at these distances in the halo typically achieve $\sim$25\% uncertainties \citep[e.g.,][]{deason12b}.

@@

The first maps of the Sgr core \citep{ibata94, ibata95} revealed
its elongated morphology and suggested tidal deformation, as would be
anticipated from its proximity to the Galaxy. Subsequent N-body models
confirmed this interpretation \citep{velazquez95} and showed that
debris from Sgr should form coherent streams of stars that encircle
the Galaxy \citep{johnston95}. Signatures of these extensive stellar
streams have since been discovered in many studies and have been
mapped across the sky in carbon stars \citep{totten98}, M giants
selected from the Two Micron All Sky Survey \citep{majewski03}, main
sequence turnoff stars from the Sloan Digital Sky Survey
\citep{belokurov06}, and RR Lyrae in the Catalina Sky Survey
\citep{drake13}.

@@

For each star, $i$,
the phase-space distance, $D_{\rm ps}$, is computed at each timestep
$t_{j}$, and the vector with the minimum phase-space distance is stored
\begin{align}
  t^*_{i} &= \argmin_{t} D_{{\rm ps},i}\\
  \bs{A}_{i} &= (\bs{q}_{i}(t^*_{i}),\bs{p}_{i}(t^*_{i})).
\end{align}
Thus, the matrix $\bs{A}_{ik}$ contains these minimum phase-space distance vectors for each star,
where $k\in[1,6]$. Intuitively, the variance of this distribution
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars: in an incorrect potential, small differences in the
orbits of the stars relative to the orbit of the progenitor cause them
to spread out in this relative, normalized phase space. Thus, the
\emph{generalized variance} of the distribution, computed for a given
set of potential parameters, $\bs{\theta}$, is a natural choice for
the scalar objective function used in constraining the potential of
the Milky Way,
\begin{equation}
  f(\bs{\theta})=\det(\mathrm{Cov}( \bs{A}_{ik})).
\end{equation}

\subsection{Application to Simulated Data}
The LM10 simulation data (see Section \ref{sec:sgr}) is a perfect
test-bed for evaluating the effectiveness of this method. We start by
extracting both particle data and the satellite orbital parameters
from the present-day snapshot of the simulation
data.\footnote{\url{www.astro.virginia.edu/~srm4n/Sgr/data.html}} We then
`observe' a sample of 100 stars from the first leading and trailing
wraps of the stream: the radial velocity errors are drawn from a
Gaussian $\epsilon_{rv} \sim \mathcal{N}(\mu=0,\sigma=10~{\rm km/s})$,
the tangential velocity errors depend on the distance to the star and
are computed from the expected Gaia error curve (shown in
Figure~\ref{fig:gaia_errors}), and the distance errors on a star of distance
$D$ are $\epsilon_{dist} \sim \mathcal{N}(0,0.02\times
D)$. Figure~\ref{fig:variance6d} shows all two-dimensional projections of the six-dimensional minimum
phase-space distance distribution for the observed particles in the
correct and an incorrect potential: it is clear that the
variance of the distribution increases in the wrong potential. The 
generalized variance defines a convex hypersurface over which we
optimize four of the six logarithmic potential parameters:
$v_{circ}$, $\phi$, $q_1$, and $q_z$. We fix $q_2$ and $R_c$, which are degenerate with combinations of the other parameters. Figure~\ref{fig:} shows one-dimensional slices of the objective function
produced by varying each of the potential parameters by
$\pm20\%$ around the true values and holding all others fixed.

[Bootstrap]

@@

\section{Discussion and future prospects}

Many orbit fitting methods for measuring the Galactic potential assume that the stars in a given stream are on the same
orbit. \cite{binney08} show that this assumption leads to XXX\% biases in the 
potential parameters, even without realistic observational errors on
the kinematic data. The \emph{Rewinder} is
free of this assumption and instead \emph{relies} on the fact that
each star is on a different orbit, and is thus an independent potentiometer: the best potential is the one that gets the
most of these orbits to intersect the progenitor
over the history of interaction with the system. The strength of this method is in the simplicity: it requires only a rough estimate of the satellite mass $m_{\rm sat}$ combined with backwards integration of orbits. There are no assumptions made about, or N-body modeling of, the internal distribution of satellite stars.

[TODO: this paragraph needs work]
There are two other apparent strengths of the method which require further investigation.
\emph{The Rewinder} is applicable to any debris that is known to come from a single object, and not restricted only to the very coldest tidal streams. For example, in principle the
method could applied to the vast debris {\it clouds} that have been discovered subtending large solid angles \citep[e.g. the Triangulum Andromeda
and Hercules-Aquila clouds]{rochapinto04,belokurov06}, or even stars that have only associations in orbital properties and do not form a coherent spatial structure
\citep[such as the angular momentum groupings in local giants found by][]{helmi99}. Second, it appears trivial to combine constraints from multiple debris systems at once by simply integrating all debris from several satellites simultaneously with $D_{\rm ps}$ defined appropriately for each star separately.

[present the limitations]
The limitations of the method also need to be understood. Firstly, for the RR Lyraes associated with the very coldest streams \citep[e.g. the globular clusters Pal5
and GD1][]{odenkirchen02,koposov10} the measurement errors are unlikely to be small enough for the
the differences in orbital properties to between debris and satellite to be detected. [What will this cause? Maybe biases? One more sentence here]. Second, the method requires specifying a functional form for the potential. [...]. Lastly, the current version of the algorithm relies on accurate knowledge of the phase-space position of the parent satellite, which may not be available in some cases \citep[for example, for the Orphan Stream][]{belokurov07}.

[explain how some of them are handled automatically by the forthcoming machinery]
The objective function presented above is an approximation to a more general likelihood function, described in detail in forthcoming work (Price-Whelan \& Johnston in prep.). The results shown in sections xx and yy are meant to illustrate the strength and concept of the method.

Given sufficient data, position and motion of the original host could be solved for as additional free parameters,
but the effectiveness of this approach has yet to be investigated.

\section{Summary}

This paper outlines an algorithm to measure the Galactic potential that takes maximal advantage of a combination of possible data from the Spitzer and Gaia satellite
missions that, when combined,  promise full phase-space measurements of RR Lyrae stars around our Galaxy.
When applied to a synthetic data set generated from debris in an N-body simulation of the destruction of Sgr the method recovers the shape and orientation of the dark
matter halo component of the potential with ??? accuracy using a sample of ??? stars.
This success provides strong motivation for: (i) further theoretical work to investigate the power of multiple debris structures with different morphologies and orientations in
building a comprehensive picture of the Galactic mass distribution; and (ii) a  Spitzer survey of RR Lyrae stars in debris structures around the Milky Way as a ``legacy''
sample to combine with near-future Gaia data.

>>>>

The Milky Way is the best candidate  for dissecting a dark matter halo in great detail.
Our proximity means that we can use individual stars as tracer populations and hence build much larger samples that probe deeper into the halo than the globular cluster
and planetary nebula studies of external galaxies \citep[e.g.][]{mendez01,cote03}.
Moreover, our internal perspective allows us to build a three-dimensional picture of the mass distribution rather than measuring its two-dimensional projection.
%Hence we can hope to reconstruct not only the density, but also the shape and orientation of the dark matter halo as function of radius well beyond the disk.
For example, to date thousands of blue-horizontal-branch stars selected from the Sloan Digital Sky Survey (SDSS) have been used to probe mass out to tens of
kpc \citep[SDSS, see][]{deason12a,kafle12}, and estimates with combined tracers extend to 150kpc \citep{deason12b}.
\citet{loebman12} demonstrates how even larger stellar samples from current (i.e. SDSS) and future  (Gaia and LSST) stellar surveys will also be sensitive to any
asphericity of the halo.

@@

In this paper we re-examine and update  another approach to using tidal debris as a potential measure \citep[originially proposed by][]{johnston99a}  in the context of current and near-future observational capabilities, and illustrates its power with the example of Sgr.
In Section \ref{sec:context} we outline the observational prospects and Sgr properties that motivated this re-examination.
In Section \ref{sec:method} we present the updated potential measure and test it with synthetic observations of simulated Sgr debris.
In Section \ref{sec:discussion} we highlight the advantages and shortcomings of this method.
We conclude in Section \ref{sec:conclusion}.

@@

\section{Context and motivation} \label{sec:context}
The method presented in Section \ref{sec:method} takes advantage of
three distinct developments in the observations of tidal streams: (i)
the demonstration of a technique for deriving distances to individual
RR Lyrae stars with 2\% accuracies (Section \ref{sec:spitzer}); (ii)
the prospect of proper motion measurements of the same stars with
$\sim$10~$\mu$as/yr precision (Section \ref{sec:gaia}); and (iii) the
tracing of debris associated with Sgr around the entire Galaxy
(Section \ref{sec:sgr})

@@

These studies suggest that observations of RR Lyrae variables by
NASA's Spitzer mission could provide us with a stellar distance
indicator of unprecedented accuracy and precision.
\citet{madore12}
For example, using 5 stars with trigonometric parallaxes
  measured by Hubble \citep{benedict11}, \citet{madore12} measured  
  the dispersion in the
mid-IR PL relation -- at Spitzer wavelengths --  to be of order
$\sim$0.03~mag. This implies that it is possible to determine
individual distances that are good to better than $2\%$ for RR
Lyrae stars out to $\sim$60~kpc (the limit of Spitzer's ability to
detect and measure RR Lyrae variables, for modest integration
times). Distance measurements of Blue Horizontal Branch and Blue Straggler stars at these distances in the halo typically achieve $\sim$25\% uncertainties \citep[e.g.,][]{deason12b}.

@@

The first maps of the Sgr core \citep{ibata94, ibata95} revealed
its elongated morphology and suggested tidal deformation, as would be
anticipated from its proximity to the Galaxy. Subsequent N-body models
confirmed this interpretation \citep{velazquez95} and showed that
debris from Sgr should form coherent streams of stars that encircle
the Galaxy \citep{johnston95}. Signatures of these extensive stellar
streams have since been discovered in many studies and have been
mapped across the sky in carbon stars \citep{totten98}, M giants
selected from the Two Micron All Sky Survey \citep{majewski03}, main
sequence turnoff stars from the Sloan Digital Sky Survey
\citep{belokurov06}, and RR Lyrae in the Catalina Sky Survey
\citep{drake13}. More recently, apogalactic debris has been traced to large distances using a combination of blue horizontal branch, main-sequence turn-off, and red giant stars from the SDSS \citep{belokurov13}.

@@

For each star, $i$,
the phase-space distance, $D_{\rm ps}$, is computed at each timestep
$t_{j}$, and the vector with the minimum phase-space distance is stored
\begin{align}
  t^*_{i} &= \argmin_{t} D_{{\rm ps},i}\\
  \bs{A}_{i} &= (\bs{q}_{i}(t^*_{i}),\bs{p}_{i}(t^*_{i})).
\end{align}
Thus, the matrix $\bs{A}_{ik}$ contains these minimum phase-space distance vectors for each star,
where $k\in[1,6]$. Intuitively, the variance of this distribution
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars: in an incorrect potential, small differences in the
orbits of the stars relative to the orbit of the progenitor cause them
to spread out in this relative, normalized phase space. Thus, the
\emph{generalized variance} of the distribution, computed for a given
set of potential parameters, $\bs{\theta}$, is a natural choice for
the scalar objective function used in constraining the potential of
the Milky Way. If
\begin{equation}
  f(\bs{\theta})=\log \det(\mathrm{Cov}( \bs{A}_{ik})).
\end{equation}

[Still working on this section...need to figure out best way to express what I'm trying to say.] Though this is not explicitly a \emph{likelihood}, optimizing over this statistic is a very simple approximation to performing a maximum-likelihood optimization. Assuming the samples in $\bs{A}_{ik}$ are approximately normal-distributed, the variance is closely related to the Shannon entropy \citep{shannon1948}, $H$,  
\begin{equation}
  H \propto \log \sigma .
\end{equation}
Minimizing the entropy is then equivalent to minimizing the loss of \emph{information} for a given choice of parameters, $\bs{\theta}$, given the ``data,'' $\bs{A}_{ik}$, and is thus related to maximum-likelihood estimation \citep[see][for more detail]{mackay2002}.

\subsection{Application to Simulated Data} \label{sec:results}
The LM10 simulation data (see Section \ref{sec:sgr}) is a perfect
test-bed for evaluating the effectiveness of this method. We start by
extracting both particle data and the satellite orbital parameters
from the present-day snapshot of the simulation
data.\footnote{\url{www.astro.virginia.edu/~srm4n/Sgr/data.html}} We then
`observe' a sample of 100 stars from the first leading and trailing
wraps of the stream: the radial velocity errors are drawn from a
Gaussian $\epsilon_{rv} \sim \mathcal{N}(\mu=0,\sigma=10~{\rm km/s})$,
the tangential velocity errors depend on the distance to the star and
are computed from the expected Gaia error curve (shown in
Figure~\ref{fig:gaia_errors}), and the distance errors on a star of distance
$D$ are $\epsilon_{dist} \sim \mathcal{N}(0,0.02\times
D)$. Figure~\ref{fig:variance6d} shows all two-dimensional projections of the six-dimensional minimum
phase-space distance distribution for the observed particles in the
correct and an incorrect potential: it is clear that the
variance of the distribution increases in the wrong potential. The 
generalized variance defines a convex hypersurface over which we
optimize four of the six logarithmic potential parameters:
$v_{circ}$, $\phi$, $q_1$, and $q_z$. We fix $q_2$ and $R_c$, which are degenerate with combinations of the other parameters. Figure~\ref{fig:} shows one-dimensional slices of the objective function
produced by varying each of the potential parameters by
$\pm10\%$ around the true values and holding all others fixed.

In anticipation of extending the above method to include a true likelihood function, we use a parallelized Markov Chain Monte Carlo (MCMC) algorithm \citep{foremanmackey2013} to sample from our objective function.\footnote{Though MCMC is typically not an efficient optimization tool, the stochasticity and easy parallelization of the algorithm mean it easily outperforms other optimizers on this problem, where the objective function is noisy and expensive to compute.} We use the median value of the samples as a point estimate for the potential parameters. To assess the uncertainty in the derived halo parameters, we bootstrap resample our 100 stars 100 times and estimate the potential parameters with each sample. Figure~\ref{fig:bootstrap} shows the recovered parameters for each bootstrapped sample and demonstrates the power of the combined Spitzer and Gaia data: a moderately sized sample of RR Lyrae alone places strong constraints on the shape and mass of the Milky Way potential.

@@

\section{Discussion and future prospects} 
\label{sec:discussion}

Many orbit fitting methods for measuring the Galactic potential assume that the stars in a given stream are on the same
orbit. \cite{sanders13a} show that this assumption can lead to nearly 100\% errors in the 
estimated potential parameters, even without realistic observational errors on
the kinematic data. The \emph{Rewinder} is
free of this assumption and instead \emph{relies} on the fact that
each star is on a different orbit, and is thus an independent potentiometer; the best potential is the one that gets the
most of these orbits to intersect the progenitor
over the history of interaction with the system.

The strengths of this method stem from the simplicity: it requires only a rough estimate of the satellite mass $m_{\rm sat}$ combined with backwards integration of orbits. There are no assumptions made about, or N-body modeling of, the internal distribution of satellite stars. Thus, \emph{The Rewinder} is applicable to any debris that is known to come from a single object, and not restricted only to the very coldest tidal streams. In principle, the
method could applied to the vast debris {\it clouds} that have been discovered over large areas of the sky \citep[e.g. the Triangulum Andromeda
and Hercules-Aquila clouds]{rochapinto04,belokurov06}, or even stars that have only associations in orbital properties and do not form a coherent spatial structure
\citep[such as the angular momentum groupings in local giants found by][]{helmi99}. The method trivially extends to combining constraints from multiple debris systems at once by simply integrating all debris from several satellites simultaneously, with $D_{\rm ps}$ defined appropriately for each star.

It is also important to characterize the the limitations of this method. Firstly, the measurement errors for RR Lyraes associated with the very coldest streams \citep[e.g. the globular clusters Pal5 and GD1][]{odenkirchen02,koposov10} will likely be too large to resolve the minute differences in orbital properties between the debris and satellite to be detected. Second, the current version of the algorithm relies on accurate knowledge of the current position and velocity of the parent satellite, which may not be available in some cases \citep[for example, the Orphan Stream;][]{belokurov07}. Lastly, the present prescription treats the progenitor as a point-mass during orbit integration. Second-order effects in orbit evolution (e.g., dynamical friction) are currently ignored, leading to minor biases in the estimated potential parameters (for example, XXX in Figure~\ref{fig:bootstrap}).

The objective function presented above is an approximation to a more general likelihood function, described in depth in forthcoming work (Price-Whelan \& Johnston in prep.). The results shown in section \ref{sec:results} are meant to illustrate the strength and concept behind the method, but [...] [I want to say one more sentence about the upcoming work, and possibly how it will solve some of the limitations presented above.]

%With a justified probabilistic framework, we will also be able to provide iven sufficient data, position and motion of the original host could be solved for as additional free parameters, but the effectiveness of this approach has yet to be investigated.

\section{Conclusions}
\label{sec:conclusion}

This paper presents an algorithm for measuring the Galactic potential that anticipates combined data from the Spitzer and Gaia satellite missions that promise precise, full phase-space measurements of RR Lyrae stars in the halo of our Galaxy. When applied to a sample of 100 stars -- with realistic observational errors -- sampled from the \cite{law10} N-body simulation of the destruction of the Sgr dwarf satellite, the method recovers the shape and orientation of the dark matter potential XXX. This success provides strong motivation for 1) further theoretical work to [...], 2) an investigation of the power of using multiple, complimentary debris structures (e.g., with different morphologies and orientations) in building a comprehensive picture of the Galactic mass distribution, and 3) a Spitzer survey of RR Lyrae stars in debris structures around the Milky Way to get precise distances to combine with near-future Gaia velocity data.
